---
title: "Practice Exam 1"
subtitle: "Practice Exam, and general R practice"
author: Dee Ruttenberg 
date: Feb. 11, 2026
output: html_document
---

# Practice Exam

1. You run the following code
```{r, message = FALSE}
sum = 0 
list = c(2:5)
for(i in list){
  sum = sum + i
}
```

(a) What would be the output of `print(list)`? 

(b) When this function finishes running, what is `sum` equal to?

2. You run the following code

```{r, message=FALSE}
x = 10
y = 5
while(x>y){
  x = x-2
}
```

(a) When you finish running this code, what will x equal?

(b) What will y equal?

(c) Consider this very slightly different code (remember, != means 'not equal').  What is the problem with it?

```{r, eval = FALSE}
x = 10
y = 5
while(x!=y){
  x = x-2
}
```

3. Consider the following function, which takes the sum of a given number of pennies, nickels, dimes, and quarters:

```{r, message=FALSE}
coin_sum = function(p, n, d, q){
  c = p + 5*n + 10*d + 25*q
  if(c >= 100){
    return(c/100)
  }
  else{
    return (c)
  }
}
```

(a) Calculate `coin_sum(1,1,1,1)`

(b) Calculate `coin_sum(0,0,0,4)`

(c) What does the variable `n` mean in the function?

(d) Explain the purpose of the conditional (`if c>=100...`) in human terms.  Why might we have it?

